---
title: "Assigning participant identifiers"
author: "Candice C. Morey"
date: "1/9/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Assigning participant identifiers

### Always unique!

Never re-use a participant number in the same project. Even if you know data will be invalid (e.g., test participant, doesn't meet inclusion criteria). This can cause confusion later, or can over-write files that you may not want to be overwritten.

Vignette:

Always use the same participant number for the same person (e.g., if you are collecting data at multiple time points from the same person)

*Vignette:* Lewis Montgomery's ambitious MSc project involved collecting data from his participants in two sessions, because he required a lot of trials on each of four task combinations. Participants ran two sessions each, performing two of the combinations in each session. Eventually though, all of these data needed to be analyzed together. After all, that is the advantage of going to the effort to get the same data in each participant -  you get to do a within-participant analysis. Lewis ran each session with a unique identifier. This certainly prevented any data from being accidentally overwritten, but it meant that when it was time to analyze the data, Lewis needed a way to code which participants were actually the same people. There is no elegant way to do this, because there was no regularity between the ids. Below is the code we used to assign new participant ids to both of the data sets from each participant.  

```{r, eval=F}

# Also, need to link subject numbers. Lewis ran same participants on unique id numbers in different sessions. Table of links
# 1=29
d$Subject[d$Subject==1 | d$Subject==29]=129
# 2=19
d$Subject[d$Subject==2 | d$Subject==19]=219
# 3=16
d$Subject[d$Subject==3 | d$Subject==16]=316
# 4=25
d$Subject[d$Subject==4 | d$Subject==25]=425
# 5=32
d$Subject[d$Subject==5 | d$Subject==32]=532
# 6=13
d$Subject[d$Subject==6 | d$Subject==13]=613
# 7=14
d$Subject[d$Subject==7 | d$Subject==14]=714
# 8=24
d$Subject[d$Subject==8 | d$Subject==24]=824
# 9=15
d$Subject[d$Subject==9 | d$Subject==15]=915
# 10=17
d$Subject[d$Subject==10 | d$Subject==17]=1017
# 11=39
d$Subject[d$Subject==11 | d$Subject==39]=1139
# 18=23
d$Subject[d$Subject==18 | d$Subject==23]=1823
# 22=36
d$Subject[d$Subject==22 | d$Subject==36]=2236
# 26=49
d$Subject[d$Subject==26 | d$Subject==49]=2649
# 27=31
d$Subject[d$Subject==27 | d$Subject==31]=2731
# 28=60
d$Subject[d$Subject==28 | d$Subject==60]=2860
# 30=42
d$Subject[d$Subject==30 | d$Subject==42]=3042
# 33=54
d$Subject[d$Subject==33 | d$Subject==54]=3354
# 34=63
d$Subject[d$Subject==34 | d$Subject==63]=3463
# 35=40
d$Subject[d$Subject==35 | d$Subject==40]=3540
# 37=46
d$Subject[d$Subject==37 | d$Subject==46]=3746
# 38=48
d$Subject[d$Subject==38 | d$Subject==48]=3848
# 41=65
d$Subject[d$Subject==41 | d$Subject==65]=4165
# 43=59
d$Subject[d$Subject==43 | d$Subject==59]=4359
# 44=75
d$Subject[d$Subject==44 | d$Subject==75]=4475
# 45=52
d$Subject[d$Subject==45 | d$Subject==52]=4552
# 47=78
d$Subject[d$Subject==47 | d$Subject==78]=4778
# 50=57
d$Subject[d$Subject==50 | d$Subject==57]=5057
# 51=69
d$Subject[d$Subject==51 | d$Subject==69]=5169
# 53=55
d$Subject[d$Subject==53 | d$Subject==55]=5355
# 56=66
d$Subject[d$Subject==56 | d$Subject==66]=5666
# 58=67
d$Subject[d$Subject==58 | d$Subject==67]=5867
# 61=64
d$Subject[d$Subject==61 | d$Subject==64]=6164
# 62=71
d$Subject[d$Subject==62 | d$Subject==71]=6271
# 68=74
d$Subject[d$Subject==68 | d$Subject==74]=6874
# 70=72
d$Subject[d$Subject==70 | d$Subject==72]=7072
# 73=76
d$Subject[d$Subject==73 | d$Subject==76]=7376

```

While this code solved the immediate problem, it was a clunky solution that could have introduced many new errors that would be difficult to detect. What if Lewis lost his notes indicating which participant ids should be paired before running the analysis? What if those notes contained discrepancies he could not sort out (e.g., writing down the same id accidentally for two participants)? 

Furthermore, the inelegant coding solution needed to manage this is also prone to human error. Mis-typing an id numbers could mean creation of a bogus pairing, accidentally assigning three sessions to one person, etc. After doing this, we had to check carefully that all resulting ids had the right number of sessions, trials, and conditions. 

In the end, this worked out fine. But work and hassle would have been avoided if each participant was given a consistent id to start with, or alternatively if an id system was used that resulted in unique session ids that were related in some regular way. This problem would of course have been compounded if the design had been more complicated (e.g., what if there were > 2 sessions?) and as the sample size increased (imagine sorting this for 150 participants instead of ~30).


### Always arbitrary, anonymous

The system you choose for assigning participant numbers should not preserve any links between the participant's identity in real-life and their identity as a unique contributor to your data set. Ids should not be derived from any university-linked identifier, address, or birthdate. They should not include the date or time of the session. Typical ways to assign ids are to choose a starting number and increment upward with each new participant.  

If you are running a multi-experiment project, it is a good idea to segregate the participant ids of subsequent experiments in some way that makes them obviously distinct from previous experiments in the sequence (e.g., experiment 1 starts from 100, experiment 2 from 200, etc.). This way, if you ever combine data to perform a multi-experiment analysis, you still have unique participant ids. 
